Supplementary Material

### 2024-03-21 article/paper_drought_Chile2023.qmd
update article

  docx:

@@

@@

Human-induced greenhouse gas emissions have increased the frequency and/or intensity of drought as a result of global warming, according to the sixth assessment report (AR6) of the Intergovernmental Panel on Climate Change (IPCC) [@IPCC2023]. The evidence supporting this claim has been strengthened since AR5 [@IPCC2013]. Recent studies, however, have produced contrasting findings, suggesting that drought has not exhibited a significant trend over the past forty years. [@Vicente-Serrano2022; @Kogan2020]. @Vicente-Serrano2022 analyzed the meteorological drought trend on a global scale, finding that only in a few regions has there been an increase in the severity of drought. Moreover, they attribute the increase in droughts over the past forty years solely to an increase in atmospheric evaporative demand (AED), which in turn enhances vegetation water demand, with important implications for agricultural and ecological droughts. Also, they state that "the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change". Similarly, @Kogan2020 analyzed the drought trend using vegetation health methods, finding that for the globe, hemispheres, and main grain-producing countries, drought has not expanded or intensified for the last 38 years. Further, the @IPCC2021 suggests that there is a high degree of confidence that rising temperatures will increase the extent, frequency, and severity of droughts. Also, AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C. To better evaluate the impact of drought trends on ecosystems, assessments are needed that relate meteorological and soil moisture variables to their effects on vegetation.

@@

## Short- to long-term drought

@@

The drought indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months at a monthly frequency for 1981–2023 in order to be used for short- to long-term evaluation of drought. In the case of the proxy of vegetation productivity (zcNDVI) it was calculated for a time scale of six months at monthly frequency for 2000–2023. For zcNDVI, we test time scales of 1, 3, 6, and 12 months; we choose to use six months because that shows a more stable representation of vegetation productivity.

@@

```{r fig-zcNDVI_var}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "(a) Map of the linear trend of the index zcNDVI for 2000–2023. Greener colors indicate a positive trend; reder colors correspond to a negative trend and a decrease in vegetation productivity. Grey colors indicate either no vegetation or a change in land cover type for 2001–2022. (b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile. Each horizontal panel corresponds to a macrozone from 'Norte Grande' to 'Austral'."
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png")
```

@@

@fig-corTimeScale shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between different drought indicators and zcNDVI over time scales (1, 3, 6, 12, 24, and 36 months). The spatial variation of time scales reached per index is mostly for time scales above 12 months. In the case of SSI, the predominant scales are 6 and 12 months. For all indices, to the north, the time scales are higher and diminish toward the south until the south part of "Austral," where they increase. In @fig-corPerson, the map of Pearson correlation values (r) is shown. The EDDI reached correlations above 0.5 between "Norte Chico" and "Sur." The correlation changes from negative to positive toward the Andes Mountains and to the sea, just as in the northern part of "Austral." The SPI and SPEI have similar results, with the higher values in "Norte Chico" and "Centro" being higher than 0.6. Following a similar spatial pattern as EDDI but with an opposite sign. The SSI showed to be the index that has a major spatial extension with a higher correlation. It has a similar correlation to SPI and SPEI for "Norte Chico" and "Sur,"  but has improvements for "Sur."

@@

## Drought trend and attribution to land cover change

@@

This shows that the main cause of the hydrological drought in Chile was a steady drop in water supply made worse by an increase in AED, but it seems that in zones most affected by drought, the main cause is not an increase in water demand due to an intensification of cultivated land (e.g., an increase in irrigated crops). North-central Chile has experienced a decline in vegetation productivity across land cover types, which is primarily attributable to variations in water supply and soil moisture. An increase in water demand, such as an increase in the surface area of irrigated crops, could strengthen this trend.

## Impact of short- to long-term droughts on land cover types

We found that the 12-month soil moisture deficit most affects the productivity of vegetation in all land cover types along Chile. In a study in the Yangtze River Basin in China, @Jiang2020 analyzed the impact of drought on vegetation using the SPEI and the Enahanced Vegetation Index (EVI). They found that cropland was more sensitive to drought than grassland, showing that cropland responds strongly to short- and medium-term drought (< SPEI-6). In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro." In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016;@Potopova2015;@Dai2020;@Rhee2010]. Short-term precipitation deficits impact soil water, and thus less water is available for plant growth. However, we found that in "Norte Chico," an SPI-36 and SPEI-12 had a higher impact, which are associated with hydrological drought (long-term), and in "Centro," an SPI-12 and SPEI-12. Thus, we attribute this impact to the hydrological drought that has decreased groundwater storage [@Taucare2024], which in turn is impacted by long-term deficits, and consequently, the vegetation is more dependent on groundwater. In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared. The possible reason for this is that the most resistant types, forest and grassland, predominate south of "Centro." Also, drought episodes have been less frequent and intense. The drought episodes have had a lower impact on water availability for vegetation.

@@

The main external factors that affect biomass production by vegetation are actual evapotranspiration and soil moisture, and the rate of ET in turn depends on the availability of water storage in the root zone. Thus, soil moisture plays a key role in land carbon uptake and, consequently, in the production of biomass [@Humphrey2021]. Moreover, @Zhang2022 indicate there is a bidirectional causality between soil moisture and vegetation productivity. Lastly, some studies have redefined agricultural drought as soil moisture drought from a hydrological perspective [@Loon2016;@Samaniego2018]. Even though soil moisture is the external factor most determinant of vegetation biomass, there are multiple internal factors, such as species, physiological characteristics, and plant hydraulics, that would affect vegetation productivity. Because of that, we believe that agricultural drought, referring to the drought that impacts vegetation productivity, is the most proper term, as originally defined by @Wilhite1985.

@@

# Supplementary material {-}

>>>>

  elsevier-docx:

@@

    number-sections: true

@@

Human-induced greenhouse gas emissions have increased the frequency and/or intensity of drought as a result of global warming, according to the sixth assessment report (AR6) of the Intergovernmental Panel on Climate Change (IPCC) [@IPCC2023]. The evidence supporting this claim has been strengthened since AR5 [@IPCC2013]. Recent studies, however, have produced contrasting findings, suggesting that drought has not exhibited a significant trend over the past forty years. [@Vicente-Serrano2022; @Kogan2020]. @Vicente-Serrano2022 analyzed the meteorological drought trend on a global scale, finding that only in a few regions has there been an increase in the severity of drought. Moreover, they attribute the increase in droughts over the past forty years solely to an increase in atmospheric evaporative demand (AED), which in turn enhances vegetation water demand, with important implications for agricultural and ecological droughts. Also, they state that "the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change". Similarly, @Kogan2020 analyzed the drought trend using vegetation health methods, finding that for the globe, hemispheres, and main grain-producing countries, drought has not expanded or intensified for the last 38 years. Further, @IPCC2021 suggests that there is a high degree of confidence that rising temperatures will increase the extent, frequency, and severity of droughts. Also, AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C. To better evaluate the impact of drought trends on ecosystems, assessments are needed that relate meteorological and soil moisture variables to their effects on vegetation.

@@

## Short- to long-term drought trends

@@

The drought indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months at a monthly frequency for 1981–2023 in order to be used for short- to long-term evaluation of drought. In the case of the proxy of vegetation productivity (zcNDVI), it was calculated for a time scale of six months at monthly frequency for 2000–2023. For zcNDVI, we test time scales of 1, 3, 6, and 12 months in December and its correlation with net primary production (NPP) obtained from the MOD17A3HGF product from MODIS. We choose to use six months because r-squared with NPP increases from one to six months and from six to 12 months has little improvement (see supplementary material in Section S5).

@@

```{r fig-zcNDVI_var}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "(a) Map of the linear trend of the index zcNDVI for 2000–2023. Greener colors indicate a positive trend; reder colors correspond to a negative trend and a decrease in vegetation productivity. Grey colors indicate either no vegetation or a change in land cover type for 2001–2022. (b) Temporal variation of zcNDVI aggregated at macrozone level within continental Chile. Each horizontal panel corresponds to a macrozone from 'Norte Grande' to 'Austral'."
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png")
```

@@

@fig-corTimeScale shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between zcNDVI and  different drought indicators over time scales of 1, 3, 6, 12, 24, and 36 months. The spatial variation of time scales reached per index is mostly for time scales above 12 months. In the case of SSI, the predominant scales are 6 and 12 months. For all indices, to the north, the time scales are higher and diminish toward the south until the south part of "Austral," where they increase. In @fig-corPerson, the map of Pearson correlation values (r) is shown. The EDDI reached correlations above 0.5 between "Norte Chico" and "Sur." The correlation changes from negative to positive toward the Andes Mountains and to the sea, just as in the northern part of "Austral." The SPI and SPEI have similar results, with the higher values in "Norte Chico" and "Centro" being higher than 0.6. Following a similar spatial pattern as EDDI but with an opposite sign. The SSI showed to be the index that has a major spatial extension with a higher correlation. It has a similar correlation to SPI and SPEI for "Norte Chico" and "Sur,"  but has improvements for "Sur."

@@

## The main drivers of drought in Chile

@@

This shows that the main cause of the hydrological drought in Chile was a steady drop in water supply made worse by an increase in AED, but it seems that in zones most affected by drought, the main cause is not an increase in vegetation water demand due to an intensification of cultivated land (e.g., an increase in irrigated crops). North-central Chile has experienced a decline in vegetation productivity across land cover types, which is primarily attributable to variations in water supply and soil moisture. An increase in water demand, such as an increase in the surface area of irrigated crops, could strengthen this trend. But it is out of the scope of this study. Future work should focus on the regions where the drought has been more severe and has a high proportion of irrigated crops to get insight on the real impact of irrigation on those zones.

## Land cover sensitivity to drought

We analyzed two main impacts of drought on land cover. First, the attribution of drought to the change in surface area per land cover type. Drought accounts for about 30% of the surface change per land cover type, with the exception of croplands. The main variables associated with these changes are the increase in AED and, in second place, the decrease in precipitation. Second, we analyzed the time series of drought indices and vegetation productivity per land cover type. In this case, the most important variables that had an impact on zcNDVI were the soil moisture deficit, followed by the precipitation deficit, and in third place, AED.

In a study in the Yangtze River Basin in China, @Jiang2020 analyzed the impact of drought on vegetation using the SPEI and the Enahanced Vegetation Index (EVI). They found that cropland was more sensitive to drought than grassland, showing that cropland responds strongly to short- and medium-term drought (< SPEI-6). In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro." In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016;@Potopova2015;@Dai2020;@Rhee2010]. Short-term precipitation deficits impact soil water, and thus less water is available for plant growth. However, we found that in "Norte Chico," an SPI-36 and SPEI-12 had a higher impact, which are associated with hydrological drought (long-term), and in "Centro," an SPI-12 and SPEI-12. Thus, we attribute this impact to the hydrological drought that has decreased groundwater storage [@Taucare2024], which in turn is impacted by long-term deficits, and consequently, the vegetation is more dependent on groundwater. In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared. The possible reason for this is that the most resistant types, forest and grassland, predominate south of "Centro." Also, drought episodes have been less frequent and intense. The drought episodes have had a lower impact on water availability for vegetation.

@@

We found that the 12-month soil moisture deficit most affects the productivity of vegetation in all land cover types along Chile. The main external factors that affect biomass production by vegetation are actual evapotranspiration and soil moisture, and the rate of ET in turn depends on the availability of water storage in the root zone. Thus, soil moisture plays a key role in land carbon uptake and, consequently, in the production of biomass [@Humphrey2021]. Moreover, @Zhang2022 indicate there is a bidirectional causality between soil moisture and vegetation productivity. Lastly, some studies have redefined agricultural drought as soil moisture drought from a hydrological perspective [@Loon2016;@Samaniego2018]. Even though soil moisture is the external factor most determinant of vegetation biomass, there are multiple internal factors, such as species, physiological characteristics, and plant hydraulics, that would affect vegetation productivity. Because of that, we believe that agricultural drought, referring to the drought that impacts vegetation productivity, is the most proper term, as originally defined by @Wilhite1985.

@@

### 2024-03-21 article/paper_drought_Chile2023.tex
update article

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

Human-induced greenhouse gas emissions have increased the frequency
and/or intensity of drought as a result of global warming, according to
the sixth assessment report (AR6) of the Intergovernmental Panel on
Climate Change (IPCC) \citep{IPCC2023}. The evidence supporting this
claim has been strengthened since AR5 \citep{IPCC2013}. Recent studies,
however, have produced contrasting findings, suggesting that drought has
not exhibited a significant trend over the past forty years.
\citep{Vicente-Serrano2022, Kogan2020}. \citet{Vicente-Serrano2022}
analyzed the meteorological drought trend on a global scale, finding
that only in a few regions has there been an increase in the severity of
drought. Moreover, they attribute the increase in droughts over the past
forty years solely to an increase in atmospheric evaporative demand
(AED), which in turn enhances vegetation water demand, with important
implications for agricultural and ecological droughts. Also, they state
that ``the increase in hydrological droughts has been primarily observed
in regions with high water demand and land cover change''. Similarly,
\citet{Kogan2020} analyzed the drought trend using vegetation health
methods, finding that for the globe, hemispheres, and main
grain-producing countries, drought has not expanded or intensified for
the last 38 years. Further, the \citet{IPCC2021} suggests that there is
a high degree of confidence that rising temperatures will increase the
extent, frequency, and severity of droughts. Also, AR6 \citep{IPCC2023}
predicts that many regions of the world will experience more severe
agricultural and ecological droughts even if global warming stabilizes
at 1.5°--2°C. To better evaluate the impact of drought trends on
ecosystems, assessments are needed that relate meteorological and soil
moisture variables to their effects on vegetation.

@@

\hypertarget{short--to-long-term-drought}{%
\subsection{Short- to long-term
drought}\label{short--to-long-term-drought}}

@@

The drought indices were calculated for time scales of 1, 3, 6, 12, 24,
and 36 months at a monthly frequency for 1981--2023 in order to be used
for short- to long-term evaluation of drought. In the case of the proxy
of vegetation productivity (zcNDVI) it was calculated for a time scale
of six months at monthly frequency for 2000--2023. For zcNDVI, we test
time scales of 1, 3, 6, and 12 months; we choose to use six months
because that shows a more stable representation of vegetation
productivity.

@@

\hypertarget{short--to-long-term-drought-1}{%
\subsection{Short- to long-term
drought}\label{short--to-long-term-drought-1}}

@@

\caption{\label{fig-zcNDVI_var}(a) Map of the linear trend of the index
zcNDVI for 2000--2023. Greener colors indicate a positive trend; reder
colors correspond to a negative trend and a decrease in vegetation
productivity. Grey colors indicate either no vegetation or a change in
land cover type for 2001--2022. (b) Temporal variation of zcNDVI-6
aggregated at macrozone level within continental Chile. Each horizontal
panel corresponds to a macrozone from `Norte Grande' to `Austral'.}

@@

Figure~\ref{fig-corTimeScale} shows the highest coefficient of
determination (r-squared, or rsq) found in the regression analysis
between different drought indicators and zcNDVI over time scales (1, 3,
6, 12, 24, and 36 months). The spatial variation of time scales reached
per index is mostly for time scales above 12 months. In the case of SSI,
the predominant scales are 6 and 12 months. For all indices, to the
north, the time scales are higher and diminish toward the south until
the south part of ``Austral,'' where they increase. In
Figure~\ref{fig-corPerson}, the map of Pearson correlation values (r) is
shown. The EDDI reached correlations above 0.5 between ``Norte Chico''
and ``Sur.'' The correlation changes from negative to positive toward
the Andes Mountains and to the sea, just as in the northern part of
``Austral.'' The SPI and SPEI have similar results, with the higher
values in ``Norte Chico'' and ``Centro'' being higher than 0.6.
Following a similar spatial pattern as EDDI but with an opposite sign.
The SSI showed to be the index that has a major spatial extension with a
higher correlation. It has a similar correlation to SPI and SPEI for
``Norte Chico'' and ``Sur,''~ but has improvements for ``Sur.''

@@

\hypertarget{drought-trend-and-attribution-to-land-cover-change}{%
\subsection{Drought trend and attribution to land cover
change}\label{drought-trend-and-attribution-to-land-cover-change}}

@@

This shows that the main cause of the hydrological drought in Chile was
a steady drop in water supply made worse by an increase in AED, but it
seems that in zones most affected by drought, the main cause is not an
increase in water demand due to an intensification of cultivated land
(e.g., an increase in irrigated crops). North-central Chile has
experienced a decline in vegetation productivity across land cover
types, which is primarily attributable to variations in water supply and
soil moisture. An increase in water demand, such as an increase in the
surface area of irrigated crops, could strengthen this trend.

\hypertarget{impact-of-short--to-long-term-droughts-on-land-cover-types}{%
\subsection{Impact of short- to long-term droughts on land cover
types}\label{impact-of-short--to-long-term-droughts-on-land-cover-types}}

We found that the 12-month soil moisture deficit most affects the
productivity of vegetation in all land cover types along Chile. In a
study in the Yangtze River Basin in China, \citet{Jiang2020} analyzed
the impact of drought on vegetation using the SPEI and the Enahanced
Vegetation Index (EVI). They found that cropland was more sensitive to
drought than grassland, showing that cropland responds strongly to
short- and medium-term drought (\textless{} SPEI-6). In our case, the
SPEI-12 was the one that most impacted the croplands in ``Norte Chico''
and ``Centro.'' In general, most studies show that croplands are most
sensitive to short-term drought (\textless{} SPI-6)
\citep{Zambrano2016, Potopova2015, Dai2020, Rhee2010}. Short-term
precipitation deficits impact soil water, and thus less water is
available for plant growth. However, we found that in ``Norte Chico,''
an SPI-36 and SPEI-12 had a higher impact, which are associated with
hydrological drought (long-term), and in ``Centro,'' an SPI-12 and
SPEI-12. Thus, we attribute this impact to the hydrological drought that
has decreased groundwater storage \citep{Taucare2024}, which in turn is
impacted by long-term deficits, and consequently, the vegetation is more
dependent on groundwater. In ``Sur'' and ``Austral,'' the correlations
between drought indices and vegetation productivity decrease, as do the
time scales that reach the maximum r-squared. The possible reason for
this is that the most resistant types, forest and grassland, predominate
south of ``Centro.'' Also, drought episodes have been less frequent and
intense. The drought episodes have had a lower impact on water
availability for vegetation.

@@

The main external factors that affect biomass production by vegetation
are actual evapotranspiration and soil moisture, and the rate of ET in
turn depends on the availability of water storage in the root zone.
Thus, soil moisture plays a key role in land carbon uptake and,
consequently, in the production of biomass \citep{Humphrey2021}.
Moreover, \citet{Zhang2022} indicate there is a bidirectional causality
between soil moisture and vegetation productivity. Lastly, some studies
have redefined agricultural drought as soil moisture drought from a
hydrological perspective \citep{Loon2016, Samaniego2018}. Even though
soil moisture is the external factor most determinant of vegetation
biomass, there are multiple internal factors, such as species,
physiological characteristics, and plant hydraulics, that would affect
vegetation productivity. Because of that, we believe that agricultural
drought, referring to the drought that impacts vegetation productivity,
is the most proper term, as originally defined by \citet{Wilhite1985}.

@@

\hypertarget{supplementary-material}{%
\section*{Supplementary material}\label{supplementary-material}}
\addcontentsline{toc}{section}{Supplementary material}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

Human-induced greenhouse gas emissions have increased the frequency
and/or intensity of drought as a result of global warming, according to
the sixth assessment report (AR6) of the Intergovernmental Panel on
Climate Change (IPCC) \citep{IPCC2023}. The evidence supporting this
claim has been strengthened since AR5 \citep{IPCC2013}. Recent studies,
however, have produced contrasting findings, suggesting that drought has
not exhibited a significant trend over the past forty years.
\citep{Vicente-Serrano2022, Kogan2020}. \citet{Vicente-Serrano2022}
analyzed the meteorological drought trend on a global scale, finding
that only in a few regions has there been an increase in the severity of
drought. Moreover, they attribute the increase in droughts over the past
forty years solely to an increase in atmospheric evaporative demand
(AED), which in turn enhances vegetation water demand, with important
implications for agricultural and ecological droughts. Also, they state
that ``the increase in hydrological droughts has been primarily observed
in regions with high water demand and land cover change''. Similarly,
\citet{Kogan2020} analyzed the drought trend using vegetation health
methods, finding that for the globe, hemispheres, and main
grain-producing countries, drought has not expanded or intensified for
the last 38 years. Further, \citet{IPCC2021} suggests that there is a
high degree of confidence that rising temperatures will increase the
extent, frequency, and severity of droughts. Also, AR6 \citep{IPCC2023}
predicts that many regions of the world will experience more severe
agricultural and ecological droughts even if global warming stabilizes
at 1.5°--2°C. To better evaluate the impact of drought trends on
ecosystems, assessments are needed that relate meteorological and soil
moisture variables to their effects on vegetation.

@@

\hypertarget{short--to-long-term-drought-trends}{%
\subsection{Short- to long-term drought
trends}\label{short--to-long-term-drought-trends}}

@@

The drought indices were calculated for time scales of 1, 3, 6, 12, 24,
and 36 months at a monthly frequency for 1981--2023 in order to be used
for short- to long-term evaluation of drought. In the case of the proxy
of vegetation productivity (zcNDVI), it was calculated for a time scale
of six months at monthly frequency for 2000--2023. For zcNDVI, we test
time scales of 1, 3, 6, and 12 months in December and its correlation
with net primary production (NPP) obtained from the MOD17A3HGF product
from MODIS. We choose to use six months because r-squared with NPP
increases from one to six months and from six to 12 months has little
improvement (see supplementary material in Section S5).

@@

\hypertarget{short--to-long-term-drought-trends-1}{%
\subsection{Short- to long-term drought
trends}\label{short--to-long-term-drought-trends-1}}

@@

\caption{\label{fig-zcNDVI_var}(a) Map of the linear trend of the index
zcNDVI for 2000--2023. Greener colors indicate a positive trend; reder
colors correspond to a negative trend and a decrease in vegetation
productivity. Grey colors indicate either no vegetation or a change in
land cover type for 2001--2022. (b) Temporal variation of zcNDVI
aggregated at macrozone level within continental Chile. Each horizontal
panel corresponds to a macrozone from `Norte Grande' to `Austral'.}

@@

Figure~\ref{fig-corTimeScale} shows the highest coefficient of
determination (r-squared, or rsq) found in the regression analysis
between zcNDVI and different drought indicators over time scales of 1,
3, 6, 12, 24, and 36 months. The spatial variation of time scales
reached per index is mostly for time scales above 12 months. In the case
of SSI, the predominant scales are 6 and 12 months. For all indices, to
the north, the time scales are higher and diminish toward the south
until the south part of ``Austral,'' where they increase. In
Figure~\ref{fig-corPerson}, the map of Pearson correlation values (r) is
shown. The EDDI reached correlations above 0.5 between ``Norte Chico''
and ``Sur.'' The correlation changes from negative to positive toward
the Andes Mountains and to the sea, just as in the northern part of
``Austral.'' The SPI and SPEI have similar results, with the higher
values in ``Norte Chico'' and ``Centro'' being higher than 0.6.
Following a similar spatial pattern as EDDI but with an opposite sign.
The SSI showed to be the index that has a major spatial extension with a
higher correlation. It has a similar correlation to SPI and SPEI for
``Norte Chico'' and ``Sur,''~ but has improvements for ``Sur.''

@@

\hypertarget{the-main-drivers-of-drought-in-chile}{%
\subsection{The main drivers of drought in
Chile}\label{the-main-drivers-of-drought-in-chile}}

@@

This shows that the main cause of the hydrological drought in Chile was
a steady drop in water supply made worse by an increase in AED, but it
seems that in zones most affected by drought, the main cause is not an
increase in vegetation water demand due to an intensification of
cultivated land (e.g., an increase in irrigated crops). North-central
Chile has experienced a decline in vegetation productivity across land
cover types, which is primarily attributable to variations in water
supply and soil moisture. An increase in water demand, such as an
increase in the surface area of irrigated crops, could strengthen this
trend. But it is out of the scope of this study. Future work should
focus on the regions where the drought has been more severe and has a
high proportion of irrigated crops to get insight on the real impact of
irrigation on those zones.

\hypertarget{land-cover-sensitivity-to-drought}{%
\subsection{Land cover sensitivity to
drought}\label{land-cover-sensitivity-to-drought}}

We analyzed two main impacts of drought on land cover. First, the
attribution of drought to the change in surface area per land cover
type. Drought accounts for about 30\% of the surface change per land
cover type, with the exception of croplands. The main variables
associated with these changes are the increase in AED and, in second
place, the decrease in precipitation. Second, we analyzed the time
series of drought indices and vegetation productivity per land cover
type. In this case, the most important variables that had an impact on
zcNDVI were the soil moisture deficit, followed by the precipitation
deficit, and in third place, AED.

In a study in the Yangtze River Basin in China, \citet{Jiang2020}
analyzed the impact of drought on vegetation using the SPEI and the
Enahanced Vegetation Index (EVI). They found that cropland was more
sensitive to drought than grassland, showing that cropland responds
strongly to short- and medium-term drought (\textless{} SPEI-6). In our
case, the SPEI-12 was the one that most impacted the croplands in
``Norte Chico'' and ``Centro.'' In general, most studies show that
croplands are most sensitive to short-term drought (\textless{} SPI-6)
\citep{Zambrano2016, Potopova2015, Dai2020, Rhee2010}. Short-term
precipitation deficits impact soil water, and thus less water is
available for plant growth. However, we found that in ``Norte Chico,''
an SPI-36 and SPEI-12 had a higher impact, which are associated with
hydrological drought (long-term), and in ``Centro,'' an SPI-12 and
SPEI-12. Thus, we attribute this impact to the hydrological drought that
has decreased groundwater storage \citep{Taucare2024}, which in turn is
impacted by long-term deficits, and consequently, the vegetation is more
dependent on groundwater. In ``Sur'' and ``Austral,'' the correlations
between drought indices and vegetation productivity decrease, as do the
time scales that reach the maximum r-squared. The possible reason for
this is that the most resistant types, forest and grassland, predominate
south of ``Centro.'' Also, drought episodes have been less frequent and
intense. The drought episodes have had a lower impact on water
availability for vegetation.

@@

We found that the 12-month soil moisture deficit most affects the
productivity of vegetation in all land cover types along Chile. The main
external factors that affect biomass production by vegetation are actual
evapotranspiration and soil moisture, and the rate of ET in turn depends
on the availability of water storage in the root zone. Thus, soil
moisture plays a key role in land carbon uptake and, consequently, in
the production of biomass \citep{Humphrey2021}. Moreover,
\citet{Zhang2022} indicate there is a bidirectional causality between
soil moisture and vegetation productivity. Lastly, some studies have
redefined agricultural drought as soil moisture drought from a
hydrological perspective \citep{Loon2016, Samaniego2018}. Even though
soil moisture is the external factor most determinant of vegetation
biomass, there are multiple internal factors, such as species,
physiological characteristics, and plant hydraulics, that would affect
vegetation productivity. Because of that, we believe that agricultural
drought, referring to the drought that impacts vegetation productivity,
is the most proper term, as originally defined by \citet{Wilhite1985}.

@@

### 2024-03-21 article/supplementary.qmd
update article

```{r}
#| echo: false
#| out-width: 75%
#| fig-cap: "Heatmap of trends in zcNDVI for 2000 to 2023 per macrozone and landcover macroclass."
knitr::include_graphics('../output/figs/heatmap_trends_zcNDVI_macro_landcover.png')
```

>>>>

```{r}
#| echo: false
#| out-width: 75%
#| fig-pos: '!ht'
#| fig-cap: "Heatmap of trends in zcNDVI for 2000 to 2023 per macrozone and landcover macroclass."
knitr::include_graphics('../output/figs/heatmap_trends_zcNDVI_macro_landcover.png')
```

# Vegetation productivity

We analyzed the correlation of zcNDVI for time scales of 1, 3, 6, and 12 months versus net primary production (NPP). Both were obtained from MODIS products. zcNDVI from MOD13A3.061 and NPP from MOD17A3HGF.061.

```{r}
#| echo: false
#| out-width: 75%
#| fig-pos: '!ht'
#| fig-cap: "Spatial variation of the r-squared values obtained from the yearly correlation of zcNDVI of 1, 3, 6, and 12 months with the net primary productivity (NPP) for continental Chile."
knitr::include_graphics('../output/figs/map_r2_NPP_vs_zcNDVI.png')
```

```{r}
#| echo: false
#| out-width: 75%
#| fig-pos: '!ht'
#| fig-cap: "A heatmap showing the r-squared values obtained from the yearly correlation of zcNDVI of 1, 3, 6, and 12 months with the net primary productivity (NPP) for continental Chile."
knitr::include_graphics('../output/figs/heatmap_r2_npp_vs_zcndvi_per_landcover.png')
```
